---
title: "SDS 390: Exploring the Statistical Implications of Study Design and Sampling Schemes"
date: "February 8, 2024"
format: 
  pdf:
    keep-tex: true
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
    geometry: 
      - left=1in
      - right=1in
      - top=1in
      - bottom=1in
---

```{r}
#| include: false

# A helper function to quickly calculate marginal proportions of interest
calc_marginal <- function(dat){
  p_copd <- mean(dat$disease == "COPD")
  p_no_copd <- mean(dat$disease == "no COPD")
  p_cannabis <- mean(dat$exposure == "CBD use")
  p_no_cannabis <- mean(dat$exposure == "no CBD use")
  return(data.frame(p_copd = p_copd, p_no_copd = p_no_copd, p_cannabis = p_cannabis, p_no_cannabis = p_no_cannabis))
}

# A helper function to quickly calculate joint proportions of interest
calc_joint <- function(dat){
  p_copd_cannabis <- mean(dat$disease == "COPD" & dat$exposure == "CBD use")
  p_no_copd_cannabis <- mean(dat$disease == "no COPD" & dat$exposure == "CBD use")
  p_copd_no_cannabis <- mean(dat$disease == "COPD" & dat$exposure == "no CBD use")
  p_no_copd_no_cannabis <- mean(dat$disease == "no COPD" & dat$exposure == "no CBD use")
  return(data.frame(p_copd_cannabis = p_copd_cannabis, p_no_copd_cannabis = p_no_copd_cannabis, p_copd_no_cannabis = p_copd_no_cannabis, p_no_copd_no_cannabis = p_no_copd_no_cannabis))
}

# A helper function to quickly calculate proportions conditional on exposure status
calc_conditional_on_exposure <- function(dat){
  p_copd_given_cannabis <- mean(dat$disease[dat$exposure == "CBD use"] == "COPD")
  p_copd_given_no_cannabis <- mean(dat$disease[dat$exposure == "no CBD use"] == "COPD")
  return(data.frame(p_copd_given_cannabis =  p_copd_given_cannabis, p_copd_given_no_cannabis = p_copd_given_no_cannabis))
}

# A helper function to quickly calculate proportions conditional on disease status
calc_conditional_on_disease <- function(dat){
  p_cannabis_given_copd <- mean(dat$exposure[dat$disease == "COPD"] == "CBD use")
  p_cannabis_given_no_copd <- mean(dat$exposure[dat$disease == "no COPD"] == "CBD use")
  return(data.frame(p_cannabis_given_copd =  p_cannabis_given_copd, p_cannabis_given_no_copd = p_cannabis_given_no_copd))
}
```

# Our Full Population

Imagine, for a moment, that we want to understand the distribution and determinants of disease within Everytown, USA, a community with a population of 30,000 individuals! We're particularly interested in studying the relationship between cannabis use and chronic obstructive pulmonary disease (COPD), a type of inflammatory lung disease that can make it difficult to breath. In this context, cannabis use represents our exposure ($X$) and COPD represents our disease ($Y$).

```{r}
# Creating the population data for Everytown, USA
pop_data <- data.frame(disease = c(rep("COPD", 2500), 
                                   rep("no COPD", 27500)),
                       exposure = c(rep("CBD use", 1500), 
                                    rep("no CBD use", 1000), 
                                    rep("CBD use", 8500), 
                                    rep("no CBD use", 19000)))
```

The contingency table summarizing the full population is shown below:

```{r}
# Displaying that data in a contingency table
table(pop_data$exposure, pop_data$disease)
```

Focusing first on the marginal distribution of disease (COPD) and exposure (cannabis use) in Everytown, we find that:

-   33.3% of all residents (10,000 individuals) use cannabis: $P(X = 1) = 0.33$
-   8.3% of all residents (2,500 individuals) have COPD: $P(Y = 1) = 0.083$

```{r}
# Calculating the overall proportion of residents who use cannabis
mean(pop_data$exposure == "CBD use")

# Calculating the overall proportion of residents who have COPD
mean(pop_data$disease == "COPD")
```

We also note that there are higher rates of COPD among cannabis users than non-users:

-   15% of all cannabis users reported a COPD diagnosis: $P(Y = 1 | X = 1) = 0.15$
-   5% of all non-users of cannabis reported COPD: $P(Y = 1 | X = 0) = 0.05$

```{r}
# Calculating proportions that are conditional on exposure status
#   margin = 1 calculates proportions within each row
table(pop_data$exposure, pop_data$disease) |> prop.table(margin = 1) 
```

Similarly, there are higher rates of cannabis use among those diagnosed with COPD than those without COPD:

-   60% of all individuals with COPD use cannabis: $P(X = 1 | Y = 1) = 0.60$
-   30.9% of all individuals without COPD use cannabis: $P(X = 1 | Y = 0) = 0.309$,

```{r}
# Calculating proportions that are conditional on disease status
#   margin = 2 calculates proportions within each row
table(pop_data$exposure, pop_data$disease) |> prop.table(margin = 2) 
```

# Sampling Observations From the Population

We don't have the time, energy, or budget to collect cannabis use data and COPD data on all members of Everytown. Suppose we decide instead to conduct a study using $n = 5000$ population members.

## Cross-Sectional Study: Taking a Population-Based Sample

In a cross-sectional study, we sample a subset of $n = 5000$ population members *without* regard to either their cannabis habits or their COPD diagnosis:

```{r}
set.seed(112358)

# Taking the population-based sample
n <- 5000
pop_sample <- pop_data[sample(1:nrow(pop_data), n, replace = FALSE), ]

# Contingency table of the study data
table(pop_sample$exposure, pop_sample$disease) |> addmargins()
```

## Cohort Study: Taking an Exposure-Based Sample

In a cohort study, we might select our sample of $n = 5000$ population members by sampling $n_1 = 2500$ cannabis users and $n_0 = 2500$ non-users and then ascertaining their COPD status:

```{r}
# Taking the exposure-based sample
n1 <- 2500
n0 <- 2500
exp_sample <- rbind(pop_data[sample(which(pop_data$exposure=="CBD use"), n1), ],
                    pop_data[sample(which(pop_data$exposure=="no CBD use", ), n0), ])


# Contingency table of the study data
table(exp_sample$exposure, exp_sample$disease) |> addmargins()
```

## Case-Control Study: Taking a Disease-Based Sample

Finally, in a case-control study, we select our sample conditional on their disease status. For example, we might select $n_1 = 1000$ individuals with COPD and $n_0 = 4000$ individuals without COPD and look backward to determine their cannabis use: 

```{r}
# Taking the disease-based sample
n1 <- 1000
n0 <- 4000
cc_sample <- rbind(pop_data[sample(which(pop_data$disease=="COPD"), n1), ],
                    pop_data[sample(which(pop_data$disease=="no COPD", ), n0), ])


# Contingency table of the study data
table(cc_sample$exposure, cc_sample$disease) |> addmargins()
```